State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-03-13 analysis/paper/paper.Rmd
made a few changes to some figures and tables.

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

#------------------------------------------------------
# Re-format res_df into long format of point ests.,
# CI's and CIin for both Models : STADEM, SCOBI
#------------------------------------------------------
res_long_df <- res_df %>%
  pivot_longer(cols = -(Scenario:Truth),
               names_to = "key",
               values_to = "point") %>%
  separate(key, into = c("Model","est"), sep = "_") %>%
  pivot_wider(names_from = "est",
              values_from = "point") %>%
  mutate(Model = recode(Model, 
                        "ISEMP" = "STADEM")) %>%
  rowwise() %>%
  mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
         cv = se / est,
         bias = est - Truth,
         rel_bias = bias / Truth,
         MSE = mean((bias)^2),
         inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
         Window = ifelse(grepl("H",Scenario),"Window_Err_High",
                         ifelse(grepl("L",Scenario),"Window_Err_Low","No Window Error")),
         Night = ifelse(grepl("N-R",Scenario),"Yes","No"))
```

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "N-R trap down",
                                "Trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position")) %>%
  column_spec(column = 2,
              width = "1in") %>%
  column_spec(column = c(3, 4, 5),
              width = "1.2in")

@@

sum_tab <- res_long_df %>%
  filter(Variable == 'Unique.Wild.Fish',
         Model == 'STADEM') %>%
  group_by(Scenario) %>%
  summarise(inCI = sum(inCI),
            nsims = n(),
            Sample_Mean = mean(est),
            `Relative bias` = mean(rel_bias),
            SE = sqrt(var(est)),
            `Mean CV` = mean(cv),
            RMSE = sqrt(mean(bias^2)),
            Coverage = inCI/nsims) %>%
  mutate(Scenario = str_replace_all(Scenario,
                                    "_", " "),
         Scenario = str_replace(Scenario,
                                "TrapDown",
                                "Trap Down")) %>%
  select(Scenario,
         `Relative bias`,
         `Mean CV`,
         RMSE,
         Coverage)

sum_tab %>%
  kable(digits = c(0, 3, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r lgr-tab}
lgr_est %>%
  mutate(prnt_val = if_else(is.na(cv),
                            as.character(prettyNum(round(mean), big.mark = ',')),
                            paste0(prettyNum(round(mean), big.mark = ","), " (", round(cv, 3), ")"))) %>%
  pivot_wider(id_cols = c("Species",
                          "Year"),
              names_from = "variable",
              values_from = "prnt_val") %>%
  select(Species, 
         Year,
         `Window counts` = win_cnt,
         Total = X.tot.new.all,
         Wild = X.tot.new.wild,
         Hatchery = X.tot.new.hatch,
         `Hatchery no-clip` = X.tot.new.hnc) %>%
  kable(booktabs = T,
        linesep = "",
        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r stadem-examp-fig, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

```{r rel-bias-lgd, fig.cap = "Boxplots of the relative bias of STADEM estimates for wild escapement across various scenarios (See Table 1)."}
# hdi_box = function(x) {
#   r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
#   # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
#   names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
#   r
# }
# 
# hdi_out = function(x) {
#   subset(x, x < hdi(x, 0.95)[1] | x > hdi(x, 0.95)[2])
#   # subset(x, x < quantile(x, probs = c(0.025)) | x > quantile(x, probs = c(0.975)))
# }

@@

@@

# get plotting data from particular year
plot_df = tibble(Species = "Chinook",
                 Year = "2014") %>%
  mutate(plot_title = paste(Species, Year)) %>%
  rowwise() %>%
  mutate(plot_data = map2(Species, 
                          Year,
                          .f = function(x, y) {
                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
                              as_tibble(rownames = 'var') %>%
                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
                                     param = ifelse(param == '', 'all', param)) %>%
                              select(var, param, week, everything()) %>%
                              left_join(stadem_list$weeklyData %>%
                                          rename(week = week_num) %>%
                                          rowwise() %>%
                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
                                          ungroup())%>%
                              mutate_at(vars(trap_est),
                                        list(~ if_else(. == Inf,
                                                       as.numeric(NA),
                                                       .))) %>%
                              select(-Species)
                            return(plot_df)
                          })) %>%
  unnest(cols = plot_data) %>%
  select(Species, Year, plot_title, everything()) %>%
  filter(param == "all")

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

#------------------------------------------------------
# Re-format res_df into long format of point ests.,
# CI's and CIin for both Models : STADEM, SCOBI
#------------------------------------------------------
res_long_df <- res_df %>%
  pivot_longer(cols = -(Scenario:Truth),
               names_to = "key",
               values_to = "point") %>%
  separate(key, into = c("Model","est"), sep = "_") %>%
  pivot_wider(names_from = "est",
              values_from = "point") %>%
  mutate(Model = recode(Model, 
                        "ISEMP" = "STADEM")) %>%
  rowwise() %>%
  mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
         cv = se / est,
         # bias = est - Truth,
         bias = Truth - est,
         rel_bias = bias / Truth,
         MSE = mean((bias)^2),
         inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
         Window = ifelse(grepl("H",Scenario),"Window_Err_High",
                         ifelse(grepl("L",Scenario),"Window_Err_Low","No Window Error")),
         Night = ifelse(grepl("N-R",Scenario),"Yes","No"))
```

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "Trap down",
                                "N-R trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position")) %>%
  column_spec(column = 2,
              width = "1in") %>%
  column_spec(column = c(3, 4, 5),
              width = "1.2in")

@@

sum_tab <- res_long_df %>%
  filter(Variable == 'Unique.Wild.Fish',
         Model == 'STADEM') %>%
  group_by(Scenario) %>%
  summarise(inCI = sum(inCI),
            nsims = n(),
            Sample_Mean = mean(est),
            `Mean bias` = mean(bias),
            `Relative bias` = mean(rel_bias),
            SE = sqrt(var(est)),
            `Mean SE` = mean(se),
            `Mean CV` = mean(cv),
            RMSE = sqrt(mean(bias^2)),
            Coverage = inCI/nsims) %>%
  mutate(Scenario = str_replace_all(Scenario,
                                    "_", " "),
         Scenario = str_replace(Scenario,
                                "TrapDown",
                                "Trap Down")) %>%
  select(Scenario,
         `Relative bias`,
         `Mean bias`,
         `Mean SE`,
         `Mean CV`,
         RMSE,
         Coverage)

sum_tab %>%
  kable(digits = c(0, 3, 1, 0, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r lgr-tab}
lgr_est %>%
  mutate(prnt_val = if_else(is.na(cv),
                            as.character(prettyNum(round(mean), big.mark = ',')),
                            paste0(prettyNum(round(mean), big.mark = ","), " (", round(cv, 3), ")"))) %>%
  pivot_wider(id_cols = c("Species",
                          "Year"),
              names_from = "variable",
              values_from = "prnt_val") %>%
  left_join(lgr_est %>%
              group_by(Species, Year) %>%
              summarise(win_bias = mean[variable == "win_cnt"] - mean[variable == "X.tot.new.all"],
                        win_rel_bias = win_bias / mean[variable == "win_cnt"])) %>%
  mutate(across(win_rel_bias,
                ~ paste(round(. * 100, 1), "%"))) %>%
  select(Species, 
         Year,
         `Window counts` = win_cnt,
         `Window bias` = win_rel_bias,
         Total = X.tot.new.all,
         Wild = X.tot.new.wild,
         Hatchery = X.tot.new.hatch,
         `Hatchery no-clip` = X.tot.new.hnc) %>%
  kable(booktabs = T,
        linesep = "",
        caption = "Window counts, the relative bias of window counts compared to estimates of total escapement, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r dag-fig, fig.height = 5, fig.cap = "Directed acyclic graph showing the STADEM model framework."}
include_graphics("../figures/ModelSchematic.png")
```

```{r stadem-examp-fig, eval = F, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

@@

```{r rel-bias-lgd-old, eval = F, fig.cap = "Boxplots of the relative bias of STADEM estimates for wild escapement across various scenarios (See Table 1)."}
# hdi_box = function(x) {
#   r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
#   # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
#   names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
#   r
# }
# 
# hdi_out = function(x) {
#   subset(x, x < hdi(x, 0.95)[1] | x > hdi(x, 0.95)[2])
#   # subset(x, x < quantile(x, probs = c(0.025)) | x > quantile(x, probs = c(0.975)))
# }

@@

```{r rel-bias-lgd, fig.cap = "Boxplots of the relative bias of window counts and STADEM estimates for total unique fish and STADEM estimates of unique wild fish across various scenarios (See Table 1)."}
rel_bias_p = res_df %>%
  mutate(across(sim,
                as.integer)) %>%
  filter(grepl('Unique', Variable)) %>%
  group_by(Scenario,
           sim) %>%
  summarise(Truth = sum(Truth),
            win_cnts = sum(SCOBI_est),
            model_est = sum(ISEMP_est),
            .groups = "drop") %>%
  tibble::add_column(Variable = "Total Unique Fish",
                     .after = "sim") %>%
  bind_rows(res_df %>%
              mutate(across(sim,
                            as.integer)) %>%
              filter(Variable == "Unique.Wild.Fish") %>%
              mutate(Variable = recode(Variable,
                                       "Unique.Wild.Fish" = "Unique Wild Fish")) %>%
              select(Scenario, sim,
                     Variable,
                     Truth,
                     model_est = ISEMP_est)) %>%
  mutate(model_bias = model_est - Truth,
         win_bias = win_cnts - Truth) %>%
  mutate(across(ends_with("bias"),
                ~ . / Truth)) %>%
  select(Scenario, sim,
         Variable,
         ends_with("bias")) %>%
  pivot_longer(cols = ends_with("bias"),
               names_to = "source",
               values_to = "bias") %>%
  mutate(source = recode(source,
                         "model_bias" = "Model",
                         "win_bias" = "Window Count")) %>%
  ggplot(aes(x = bias,
             y = fct_rev(Scenario),
             fill = source)) +
  scale_fill_grey(start = 0.5,
                  end = 0.9) +
  geom_boxplot() +
  geom_vline(xintercept = 0,
             linetype = 2) +
  theme(axis.text = element_text(size = 10)) +
  facet_wrap(~ Variable) +
  labs(x = "Relative Bias",
       y = "Scenario",
       fill = "Source") +
  coord_fixed(0.03)

@@

# get plotting data from particular year
plot_df = tibble(Species = "Chinook",
                 Year = "2014") %>%
  mutate(plot_title = paste(Species, Year)) %>%
  rowwise() %>%
  mutate(plot_data = map2(Species, 
                          Year,
                          .f = function(x, y) {
                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', x, '_', y, '.rda'))
                            # plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
                            plot_df = stadem_mod$summary[grep('^X.new', rownames(stadem_mod$summary)),] %>%  
                              as_tibble(rownames = 'var') %>%
                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
                                     param = recode(param, "tot" = "all")) %>%
                                     # param = ifelse(param == '', 'all', param)) %>%
                              select(var, param, week, everything()) %>%
                              left_join(stadem_list$weeklyData %>%
                                          rename(week = week_num) %>%
                                          rowwise() %>%
                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
                                          ungroup())%>%
                              mutate_at(vars(trap_est),
                                        list(~ if_else(. == Inf,
                                                       as.numeric(NA),
                                                       .))) %>%
                              select(-Species)
                            return(plot_df)
                          })) %>%
  unnest(cols = plot_data) %>%
  select(Species, Year, plot_title, everything()) %>%
  filter(param == "all")
